Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-03 causal-bandit/icml-2015/tech.tex
updates
started to clean up Sec 3
updates to Theorem 1

\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\Pn}[2]{\operatorname{P}_{#1}\left\{#2\right\}}

\subsection*{Truncated Importance Sampling}

Let $P$ and $Q$ be distributions on $\Omega = \set{0,1}^N$ and let $Y \in \set{0,1}$ be defined in terms of its conditional expectation given $X \in \Omega$.
\eq{
\forall x \in \Omega \qquad \EE[Y|X = x] = r(x)\,,
}
where $r:\Omega \to [0,1]$ is some unknown function.
We tackle the problem of estimating $\mu = \EE_Q r = \sum_x \Q{x} r(x)$ using
samples of $X$ and $Y$ with $X$ sampled from $P$.
Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
\eqn{
\label{eq:truncated}
Z = Y R \ind{R < B}\,,
}
which trivially satisfies $|Z| \leq B$ surely.
Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
Computing the expectation we see
\eq{
\EE_P[Z] = \mu - \beta\,.
}
where $\beta \geq 0$ is the negative bias and satisfies
\eq{
\beta \leq \Q{R \geq B}\,.
}
Bounding the variance is also straightforward.
\eq{
\Var[Z] 
\leq \EE[Z^2] 
=\sum_x \P{X = x} Z(x)^2 
\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 
\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
= V\,.
}
The bias can be controlled in terms of $V$ via Markov's inequality.
\eq{
\beta 
\leq \Q{R \geq B} 
\leq \frac{\EE_Q[R]}{B}
= \frac{V}{B}\,.
}
Finally let $\delta > 0$ be the desired confidence level and choose 
\eq{
B = \sqrt{\frac{V n}{\log\frac{1}{\delta}}}\,,
}
which implies that the negative bias satisfies
\eq{
\beta \leq \sqrt{\frac{V}{n} \log \frac{1}{\delta}}\,. 
}
Let $Z_1,\ldots,X_n$ be $n$ i.i.d.\ copies of $Z$. Then by the Bernstein inequality we have
\eq{
\P{\left|\frac{1}{n} \sum_{t=1}^n Z - \mu\right| \geq \sqrt{\frac{V}{n} \log\frac{1}{\delta}}} \leq \delta\,.
}

\subsection*{Mixing Truncate Importance Sampling}

Now we study the problem of estimating $\mu = \EE_Q \P{Y = 1|X}$ from samples drawn
from a set of sampling distributions $\operatorname{P}_1,\ldots,\operatorname{P}_k$ with $N_i$ samples
drawn of $\operatorname{P}_i$.
Let $R_i(x) = \Q{x} / \Pn{i}{x}$ and
\eq{
V_i = \sum_{x \in \Omega} \Q{X = x} \frac{\Q{X = x}}{\Pn{i}{X = x}}
}
and
\eq{
Z_i = Y R_i \ind{R_i < B_i}\,,
}
where
\eq{
B_i = V_i \sqrt{\frac{\sum_{i=1}^k \frac{N_i}{V_i}}{\log \frac{1}{\delta}}}\,,
}
which for $k = 1$ reduces to the choice of $B$ in the previous section.
Then $\EE_{\operatorname{P_i}} Z_i = \mu - \beta$ where
\eq{
0 \leq \beta_i 
\leq \frac{V_i}{B_i}
= \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_i}{V_i}} \log\frac{1}{\delta}}\,.
}
Let $\eta_i \in [0,1]$ be a mixture coefficient that minimises the variance of the mixture 
estimator, which weights samples from $\operatorname{P_i}$ by a bound on its inverse variance, which is $V_i / N_i$.
\eq{
\eta_i = \frac{\frac{N_i}{V_i}}{\sum_k \frac{N_k}{V_k}}\,.
}
Let $Z_{i,1},\ldots,Z_{i,N_i}$ be $N_i$ copies of $Z_i$ for each $i$.
Then we define the mixture estimator by
\eq{
\hat \mu = \frac{\sum_i \eta_i \sum_{t=1}^{N_i} \frac{Z_{it}}{N_i}}{\sum_i \eta_i} 
}
Now we prepare to use Bernstein's inequality to control the concentration of $\hat \mu$ about its mean.
\eq{
b 
= \max_i \frac{\frac{\eta_i B_i}{N_i}}{\sum_i \eta_i} 
= \max_i \frac{\alpha}{\sum_i \frac{N_i}{V_i}} 
= \frac{1}{\alpha \sqrt{\log\frac{1}{\delta}}}\,.
}
Therefore
\eq{
\P{\left|\hat \mu - \mu\right| \geq \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_k}{V_k}} \log \frac{1}{\delta}}} \leq \delta\,.
}

\subsection*{General Case}

Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.

\begin{algorithm}[H]
\caption{CheckValid}\label{alg:alloc}
\begin{algorithmic}
\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
\FOR{$i \in \set{o_1,\ldots,o_N}$}
\STATE Compute variances:
\eq{
\forall k \in \calP_i \text{ and } l \in \set{0,1} \qquad
V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
}
\STATE Compute variance for learning from do nothing action:
\eq{
V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
}
\STATE Compute required budget: 
\eq{
N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
} 
\ENDFOR
\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
\STATE {\bf return true}
\ELSE
\STATE {\bf return false}
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{theorem}
Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which \cref{alg:alloc} returns true. 
Let $N_{ij}$ as computed by \cref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
and computes estimates using the mixture estimator described in the previous section. Choosing $I_T = \argmax_{ij} \hat \mu_{ij}$ leads
to
\eq{
\EE[r_T] \leq \epsilon + \frac{1}{T}\,.
}
\end{theorem}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PROOF OF SIMPLE REGRET UPPER BOUND
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}

Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
are independent and permutations of the variables can be pushed to the reward function.
The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.

\begin{lemma}\label{lem:conc1}
Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
\eq{
\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
}
\end{lemma}

\begin{lemma}\label{lem:conc2}
Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE X_i = p$ and $\delta \in [0,1]$.
Then 
\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \leq 4\delta\,.
}
\end{lemma}

\begin{proof}
For $\delta \geq 1/4$ the result is trivial. Otherwise 
by Hoeffding's bound and the union bound:
\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
&\leq \sum_{t = n_0}^\infty \P{\left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \\
&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{t}{n_0} \log \frac{2}{\delta}\right) 
\leq 4\delta\,. \qedhere
}
\end{proof}

\begin{lemma}\label{lem:m_est}
Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
\eq{
\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
}
\end{lemma}

\begin{proof}
Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
\eq{
\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
}
Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
Towards this end recall that $m(\vec{q}) = \min\set{i : q_i \geq 1 / i}$.
Therefore for $i > m$ we have $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
\eq{
\frac{3}{4} 
&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
\geq \hat q_i \\
&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
\geq q_i - \sqrt{\frac{q_i}{8m}}
\geq \frac{1}{2m}\,.
}
Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m$.
For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m$ that
\eq{
\hat q_i 
\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
\leq \frac{1}{m} \left(1 + \sqrt{\frac{1}{8}}\right)
\leq \frac{3}{2m}\,.
}
Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
\end{proof}

\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
Let $\delta = m = m(\vec{q}) / N$. Then by Lemma \ref{lem:m_est} we have 
\eq{
\P{2m/3 \leq m(\vec{\hat q}) \leq 2m} \geq 1 - \delta\,.
}
Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m)$ samples.
Therefore by Hoeffding's inequality and the union bound we have
\eq{
\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{8m}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
}
For arms not in $a$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
Therefore if $a = do(X_i = j)$, then 
\eq{
\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m}\,. 
}
Therefore $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} \geq T/4m$
and by Lemma \ref{lem:conc2} we have
\eq{
\P{\sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{T}{4m} \text{ and } \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}} \leq 4\delta / N\,.
}
Therefore with probability at least $1 - 6\delta$ we have
\eq{
(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{N}{\delta}} = \epsilon\,.
}
If this occurs, then 
\eq{
\mu_{\hat a^*} \geq \hat \mu_{\hat a^*} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_{\hat a^*}] 
\leq 6\delta + \epsilon 
\leq \frac{6m}{T} + \sqrt{\frac{32m}{T} \log \frac{NT}{m}}\,, 
}
which completes the result.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LOWER BOUND
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}

\todot{Change this to Bernoulli}
Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
For each $i$ define reward function $r_i$ by
\eq{
r_0(\boldsymbol{X}) &= \frac{1}{2} &
r_i(\boldsymbol{X}) &= \begin{cases}
\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
\frac{1}{2} & \text{otherwise}\,,
\end{cases}
}
where $\epsilon > 0$ is some constant to be chosen later.
We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
on all observations over all $T$ rounds and $\mathbb E_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
\eq{
\Prz{I_T = i} + \Pri{I_T \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
}
where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
Let $T_i(T) = \sum_{t=1}^T \ind{I_t = i}$ be the total number of times the learner intervenes on variable $i$ (with any intervention). 
Then for $i \leq m$ a computation shows that 
\todot{Need to add at minimum some citations here}
\eq{
\KL(P_0, P_i) 
&\leq 4\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
%&= \frac{\epsilon^2}{2} \mathbb E_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + q_i T\right) \\
&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + \frac{T}{m}\right)\,.
}
Define set $A$ by
\eq{
A = \set{i \leq m : \mathbb E_0 T_i(T) \leq 2T / m}\,.
}
Then for $i \in A$ and choosing $\epsilon = \sqrt{2m/3T}$ we have
\eq{
\KL(P_0, P_i) \leq \frac{3T\epsilon^2}{2m} = 1\,. 
}
Now $\sum_{i=1}^m \mathbb E_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
Therefore
\eq{
\sum_{i \in A} \Pri{I_T \neq i} 
&\geq \sum_{i \in A} \exp\left(-\KL(P_0, P_i)\right) - 1\\
&\geq \frac{|A|}{e} - 1 
\geq \frac{m}{2e} - 1\,.
}
Therefore there exists an $i \in A$ such that
\eq{
\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}\,.
}
Therefore
\eq{
R_{T,i} \geq \frac{1}{2} P_i(I_T \neq i) \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{2m}{3T}}
}
as required.
